---
title: "SHIP TMAO"
author: "Sven Kleine Bardenhorst"
date: "27 5 2020"
output:
  rmdformats::material:
    highlight: kate
    thumbnails: true
    code_folding: show
    fig_width: 10
    fig_height: 6
    df_print: kable
---

```{css, echo = FALSE}
.header-panel {
    background-color: #940a21;
    min-height: 144px;
    position: relative;
    z-index: 3;
}
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load libraries, include=FALSE}
library(tidyverse)
library(haven)
library(Hmisc)
library(dplyr)
library(knitr)
library(DiagrammeR)
library(summarytools)
library(hablar)
library(labelled)
library(car)
library(xtable)
library(huxtable)
library(corrplot)
library(FactoMineR)
library(sjPlot)
library(MASS)
library(kableExtra)
library(reshape2)
library("rgl")

st_options(bootstrap.css     = FALSE,       # Already part of the theme so no need for it
           plain.ascii       = FALSE,       # One of the essential settings
           style             = "rmarkdown", # Idem.
           dfSummary.silent  = TRUE,        # Suppresses messages about temporary files
           footnote          = NA,          # Keeping the results minimalistic
           subtitle.emphasis = FALSE,       # For the vignette theme, this gives
           headings = F)       
                                           
```

```{r Load data, include=FALSE}
mydata <- read_dta(file = "ship_2019_137_md_s2t0_20200327.dta")
#summary(mydata)
glimpse(mydata)
data <- mydata %>% dplyr::select(c("zz_nr","stool_dna_tma_cutc","stool_dna_tma_cnta","stool_dna_tma_grdh","omics_tma_tmao","omics_tma_choline","omics_tma_betaine","omics_tma_carnitine","stool_dna_tma_16s_rdna"))
#dataselall <- data %>%
#        filter(stool_dna_tma_cutc < 99975 | stool_dna_tma_cutc > 99999)  %>%
#        filter(stool_dna_tma_cnta < 99975 | stool_dna_tma_cnta > 99999) %>%
#        filter(stool_dna_tma_grdh < 99975 | stool_dna_tma_grdh > 99999) %>%
#        filter(omics_tma_tmao < 99975 | omics_tma_tmao > 99999)  %>%
#        filter(omics_tma_choline < 99975 | omics_tma_choline > 99999) %>%
#        filter(omics_tma_betaine < 99975 | omics_tma_betaine > 99999) %>%
#        filter(omics_tma_carnitine < 99975 | omics_tma_carnitine > 99999) 


#data %>% filter(stool_dna_tma_16s_rdna > 99975 | stool_dna_tma_16s_rdna < 99999) %>% summarize( cutc = sum(stool_dna_tma_cutc < 99975 | stool_dna_tma_cutc > 99999))

#select based on 16S as all have it
dataselall <- data %>% filter(stool_dna_tma_16s_rdna < 99975 | stool_dna_tma_16s_rdna > 99999)
#dataselall <- data

###change special coding to NAs
sel <- names(dataselall)
dataselall[sel] <- lapply(dataselall[sel], function(x) replace(x,x %in% 99975:99999, NA))



###get Proband_IDs
probands_zz_nr <- dataselall$zz_nr ### 425 Probands

###get SHIPdata
SHIPt <- read_dta(file = "ship_2016_114_md_t0_20200206.dta")
SHIPs0 <- read_dta(file = "ship_2016_114_md_s0_20200206.dta")
SHIPs2 <- read_dta(file = "ship_2016_114_md_s2_20200206.dta")

###select for probands
P_SHIPt <- SHIPt %>% filter(zz_nr %in% probands_zz_nr) ### 374 Probands
P_SHIPs2 <- SHIPs2 %>% filter(zz_nr %in% probands_zz_nr) ### 51 Probands
P_SHIPs0 <- SHIPs0 %>% filter(zz_nr %in% probands_zz_nr) ### 51 Probands > same as s2 but sampled 5 years earlier - probably leave out as too few for longitudinal analyses

###select for key variables and bind rows
P_SHIPt_var <- P_SHIPt %>% dplyr::select(c("zz_nr","SEX_SHIP_T0","AGE_SHIP_T0",
                                    "t0_alko_02", # Alohol
                                    "t0_rau_01", # Raucher
                                    "t0_ernaehr_01", # Fleisch
                                    "t0_ernaehr_07", # Reis 
                                    "t0_ernaehr_08", # Salat / Gemüse, roh 
                                    "t0_ernaehr_09", # Gemüse, gekocht 
                                    "t0_ernaehr_11", # Vollkornbrot 
                                    "t0_ernaehr_10", # Obst
                                    "t0_ernaehr_13",
                                    #"ffs_pattern_t0",
                                    #"ffs_t0",
                                    "t0_diab_02","t0_stro_02","t0_mi_02","cca_lmean","cca_rmean"))
P_SHIPs2_var <- P_SHIPs2 %>% dplyr::select(c("zz_nr","SEX_SHIP2","AGE_SHIP2",
                                      "s2_alko_02", # Alohol
                                      "s2_rau_01", # Raucher
                                      "s2_ernaehr_01",
                                      "s2_ernaehr_07",
                                      "s2_ernaehr_08",
                                      "s2_ernaehr_09",
                                      "s2_ernaehr_11",
                                      "s2_ernaehr_10",
                                      "s2_ernaehr_13",
                                      #"ffs_pattern_s2",
                                      #"ffs_s2",
                                      "s2_diab_02","s2_stro_02","s2_mi_02","cca_lmean","cca_rmean"))

P_SHIPt_SHIP2_var <- rbind(as.matrix(P_SHIPt_var),as.matrix(P_SHIPs2_var)) %>% as_tibble()
P_SHIPt_SHIP2_var_tma <- P_SHIPt_SHIP2_var %>% left_join(dataselall)

###some cleaning
P_SHIPt_SHIP2_var_tma$t0_diab_02[P_SHIPt_SHIP2_var_tma$t0_diab_02 > 0] <- 1 # t0_diab_02 IST NUR INDIKATOR FÜR BEHANDLUNGSART
P_SHIPt_SHIP2_var_tma$t0_mi_02[P_SHIPt_SHIP2_var_tma$t0_mi_02 > 0] <- 1
P_SHIPt_SHIP2_var_tma$t0_stro_02[P_SHIPt_SHIP2_var_tma$t0_stro_02 > 0] <- 1

P_SHIPt_SHIP2_var_tma$t0_diab_02[is.na(P_SHIPt_SHIP2_var_tma$t0_diab_02)] <- 0
P_SHIPt_SHIP2_var_tma$t0_mi_02[is.na(P_SHIPt_SHIP2_var_tma$t0_mi_02)] <- 0
P_SHIPt_SHIP2_var_tma$t0_stro_02[is.na(P_SHIPt_SHIP2_var_tma$t0_stro_02)] <- 0

outliers <- function(model) {
     cooksd <- cooks.distance(model)
     plot(cooksd, pch="*", cex=2, main="Influential Obs by Cooks distance")  # plot cook's distance
     abline(h = 4*mean(cooksd, na.rm=T), col="red")  # add cutoff line
     text(x=1:length(cooksd)+1, y=cooksd-mean(cooksd), labels=ifelse(cooksd>4*mean(cooksd, na.rm=T),names(cooksd),""), col="red")    
}

```

# Explore variables

```{r}
detect_limit_cnta <- min(P_SHIPt_SHIP2_var_tma$stool_dna_tma_cnta[P_SHIPt_SHIP2_var_tma$stool_dna_tma_cnta > 0])
detect_limit_grdh <- min(P_SHIPt_SHIP2_var_tma$stool_dna_tma_grdh[P_SHIPt_SHIP2_var_tma$stool_dna_tma_grdh > 0])

data <-  P_SHIPt_SHIP2_var_tma %>%
         transmute(zz_nr = zz_nr,
                   meat = case_when(t0_ernaehr_01 == 1 ~ 4,  # Täglich Fleisch
                                    t0_ernaehr_01 == 4 ~ 3,
                                    t0_ernaehr_01 == 5 ~ 2,
                                    t0_ernaehr_01 == 6 ~ 1), # Nie oder fast nie Fleisch
                   eggs = case_when(t0_ernaehr_13 == 6 ~ 1,
                                    t0_ernaehr_13 == 5 ~ 2,
                                    t0_ernaehr_13 == 4 ~ 3,
                                    t0_ernaehr_13 == 3 ~ 4,
                                    t0_ernaehr_13 == 2 ~ 5,
                                    t0_ernaehr_13 == 1 ~ 6),
                   meat_cat = case_when(t0_ernaehr_01 == 1 ~ 1,
                                        t0_ernaehr_01 == 4 ~ 0,
                                        t0_ernaehr_01 == 5 ~ 0,
                                        t0_ernaehr_01 == 6 ~ 0),
          gender = case_when(SEX_SHIP_T0 == 2 ~ 0,  #Weiblich
                             SEX_SHIP_T0 == 1 ~ 1), # Männlich
          age = AGE_SHIP_T0,
          cutc = stool_dna_tma_cutc,
          cnta = case_when(stool_dna_tma_cnta > 0 ~ stool_dna_tma_cnta,
                           stool_dna_tma_cnta == 0 ~ detect_limit_cnta/sqrt(2)),
          grdh = case_when(stool_dna_tma_grdh > 0 ~ stool_dna_tma_grdh,
                           stool_dna_tma_grdh == 0 ~ detect_limit_grdh/sqrt(2)),
          tmao = omics_tma_tmao,
          choline = omics_tma_choline,
          betaine = omics_tma_betaine,
          carnitine = omics_tma_carnitine,
          IMT_mean = (cca_lmean+cca_rmean)/2,
          diabetes = t0_diab_02,
          stroke = t0_stro_02,
          mi = t0_mi_02,
          ID = zz_nr) %>%
          convert(fct(stroke,mi,diabetes)) %>%
          set_variable_labels(IMT_mean = "IMT_mean")

```


```{r Metabolites, echo=FALSE}
data %>%
  dplyr::select(meat,meat_cat,eggs,age,stroke,mi,diabetes,choline,betaine,carnitine,tmao,cutc,cnta,IMT_mean) %>%
  dfSummary() %>%
  print(max.tbl.width = 20,method="render")
```


## Correlations

```{r}
data_corr <- data %>%
        dplyr::select_if(is.numeric) %>% dplyr::select(-ID) %>% as.matrix() %>% cor(method="spearman",use="pairwise.complete.obs")

corrplot(data_corr,method="number")
```

# Analyses

## Direct effect of diet on IMT and TMAO

#### IMT
```{r echo=FALSE}
fit1 <- lm((IMT_mean)~meat_cat,data=data)
fit2 <- lm((IMT_mean)~meat_cat+age,data=data)
fit3 <- lm((IMT_mean)~meat_cat+age+gender,data=data)

tab_model(fit1,fit2,fit3)

#htmltools::includeHTML("IMT_output.html")

```

#### TMAO
```{r}
fit1 <- lm(log(tmao)~meat_cat,data=data)
fit2 <- lm(log(tmao)~meat_cat+age,data=data)
fit3 <- lm(log(tmao)~meat_cat+age+gender,data=data)

tab_model(fit1,fit2,fit3)

#htmltools::includeHTML("TMAO_output.html")
```

## Effect of TMAO on IMT
```{r}
fit1 <- lm(IMT_mean~log2(tmao),data=data)
fit2 <- lm(IMT_mean~log2(tmao)+age,data=data)
fit3 <- lm(IMT_mean~log2(tmao)+age+gender,data=data)

tab_model(fit1,fit2,fit3)

#htmltools::includeHTML("IMT_TMAO_output.html")
```

## Effect of bact on IMT
```{r}
fit1 <- lm(IMT_mean~log(cutc),data=data)
fit2 <- lm(IMT_mean~log(cnta),data=data)
fit3 <- lm(IMT_mean~log(grdh),data=data)

tab_model(fit1,fit2,fit3)
```



## Level 1 - Effect of diet on metabolites and microbiota

#### Effect of diet on choline
```{r}
fit1 <- lm(log(choline)~meat_cat,data=data)
fit2 <- lm(log(choline)~meat_cat+age,data=data)
fit3 <- lm(log(choline)~meat_cat+gender+age,data=data)
fit4 <- lm(log(choline)~gender+age,data=data)

tab_model(fit1,fit2,fit3,fit4)

#htmltools::includeHTML("choline_output.html")
```

#### Effect of diet on carnitine
```{r}
fit1 <- lm(log(carnitine)~meat_cat,data=data)
fit2 <- lm(log(carnitine)~meat_cat+age,data=data)
fit3 <- lm(log(carnitine)~meat_cat+gender+age,data=data)
fit4 <- lm(log(carnitine)~gender+age,data=data)

anova(fit3)

tab_model(fit1,fit2,fit3,fit4)

#htmltools::includeHTML("carnitine_output.html")
```

#### Effect of diet on betaine
```{r}
fit1 <- lm(log(betaine)~meat_cat,data=data)
fit2 <- lm(log(betaine)~meat_cat+age,data=data)
fit3 <- lm(log(betaine)~meat_cat+gender+age,data=data)
fit4 <- lm(log(betaine)~gender+age,data=data)

tab_model(fit1,fit2,fit3,fit4)

#htmltools::includeHTML("betaine_output.html")
```

#### Effect of diet on CutC
```{r}
fit1 <- lm(log(cutc)~meat_cat,data=data)
fit2 <- lm(log(cutc)~meat_cat+age,data=data)
fit3 <- lm(log(cutc)~meat_cat+age+gender,data=data)

tab_model(fit1,fit2,fit3)

#htmltools::includeHTML("cutc_output.html")
```

#### Effect of diet on CntA

Here, analyses were performed on the subset of observations with cnta levels above 0.
```{r}
fit1 <- lm(log(cnta)~meat_cat,data=data)
fit2 <- lm(log(cnta)~meat_cat+age,data=data)
fit3 <- lm(log(cnta)~meat_cat+age+gender,data=data)

tab_model(fit1,fit2,fit3)

#htmltools::includeHTML("cnta_output.html")
```

#### Effect of diet on GrdH
Here, analyses were performed on the subset of observations with grdh levels above 0.
```{r}
fit1 <- lm(log(grdh)~meat_cat,data=data)
fit2 <- lm(log(grdh)~meat_cat+age,data=data)
fit3 <- lm(log(grdh)~meat_cat+age+gender,data=data)

tab_model(fit1,fit2,fit3)

#htmltools::includeHTML("grdh_output.html")
```


## Level 2 - Effects of metabolites on TMAO levels

Aufgrund der extremen Skalen, habe ich die Prädiktoren hier skaliert.
```{r}
fit_tmao1 <- lm(log(tmao)~scale(choline),data=data)
fit_tmao2 <- lm(log(tmao)~scale(carnitine),data=data)
fit_tmao3 <- lm(log(tmao)~scale(betaine),data=data)
fit_tmao4 <- lm(log(tmao)~scale(choline)+scale(betaine)+scale(carnitine),data=data)


tab_model(fit_tmao1,fit_tmao2,fit_tmao3,fit_tmao4)

#htmltools::includeHTML("tmaom_output.html")
```

## Effect of microbiota on TMAO
```{r}
fit1 <- lm(log(tmao)~log(cutc),data=data)
fit2 <- lm(log(tmao)~log(cutc)+age,data=data)
fit3 <- lm(log(tmao)~log(cutc)+age+gender,data=data)
tab_model(fit1,fit2,fit3)
```


```{r}
fit1 <- lm(log(tmao)~log(cnta),data=data)
fit2 <- lm(log(tmao)~log(cnta)+age,data=data)
fit3 <- lm(log(tmao)~log(cnta)+age+gender,data=data)

tab_model(fit1,fit2,fit3)


```

```{r}
fit1 <- lm(log(tmao)~log(grdh),data=data)
fit2 <- lm(log(tmao)~log(grdh)+age,data=data)
fit3 <- lm(log(tmao)~log(grdh)+age+gender,data=data)

tab_model(fit1,fit2,fit3)
```

```{r}
fit1 <- lm(log(tmao)~log(cutc)+age+gender,data=data)
fit2 <- lm(log(tmao)~log(cnta),data=data)
fit3 <- lm(log(tmao)~scale(grdh),data=data)
fit4 <- lm(log(tmao)~log(cutc)+log(cnta)+log(grdh),data=data)

tab_model(fit1,fit2,fit3,fit4)

#htmltools::includeHTML("tmao_micro_output.html")
```




## Some ideas

### Principal Component Analysis

We can perform principal component analysis to find linear combinations of the gene and metabolite variables. As we can see in the second plot, PC1 mainly captures the metabolites, while PC2 captures the gene expressions. We may use these principal components as a proxy for the effect of metabolites (PC1) and bacteria (PC2) on TMAO levels.


```{r}
pca.res <- data %>%
        dplyr::select(cutc,cnta,grdh,choline,betaine,carnitine) %>%
        PCA(scale.unit = T,ncp=2)


pca_dat <- data %>%
        mutate(PC1=pca.res$ind$coord[,1],
               PC2=pca.res$ind$coord[,2])

fit1 <- lm(log(tmao)~PC1,data=pca_dat)
fit2 <- lm(log(tmao)~PC2,data=pca_dat)
fit3 <- lm(log(tmao)~PC1+PC2,data=pca_dat)
fit4 <- lm(log(tmao)~PC1*PC2,data=pca_dat)


fit5 <- lm(log(tmao)~PC1+age+gender,data=pca_dat)
fit6 <- lm(log(tmao)~PC2+age+gender,data=pca_dat)
fit7 <- lm(log(tmao)~PC1+PC2+age+gender,data=pca_dat)
fit8 <- lm(log(tmao)~PC1*PC2+age+gender,data=pca_dat)

tab_model(fit1,fit2,fit3,fit4)
tab_model(fit5,fit6,fit7,fit8)

#htmltools::includeHTML("pca_output.html")
#htmltools::includeHTML("pca_cov_output.html")
```

```{r include=FALSE}
# Alle Ernährungsvariablen ausprobieren?
# t0_ernaehr_08 - 16 

# t0_alko_02 Wie oft Alkohol in letzten 12 Monaten?

# t0_rau_01 aktuell Raucher?

# FFS_PATTERN_T0  - Ernährungsmuster nach FFS Score; 1 - ungünstig, 2 - intermediär , 3 - empfehlenswert



data <- data %>% left_join(P_SHIPt_SHIP2_var_tma)

data <- data %>%
        mutate(PC1=pca.res$ind$coord[,1],
               PC2=pca.res$ind$coord[,2])

reverse_scale <- function(x) {abs((max(x)+1)-x)}

data <- data %>% mutate(alcohol = reverse_scale(t0_alko_02),
                smoker = case_when(t0_rau_01 == 2 ~ 0,
                                   t0_rau_01 == 1 ~ 1),
                rice = reverse_scale(t0_ernaehr_07),
                salad = reverse_scale(t0_ernaehr_08),
                vegetables = reverse_scale(t0_ernaehr_09),
                full_grain = reverse_scale(t0_ernaehr_11),
                fruits = reverse_scale(t0_ernaehr_10))



```
                                   
# Association with confounders


## Univariate associations with cutc
```{r echo=FALSE, message=FALSE, warning=FALSE}
data %>%
        dplyr::select(meat_cat, smoker, fruits,rice,salad,vegetables,full_grain,eggs) %>%
        map(~lm(log(data$cutc) ~ .x,data=data)) %>%
        map(summary) %>%
        map(coef) %>%
        map_dbl(function(x) x[[2,4]]) %>%
        tidy %>%
        rename(p_value = x) %>%
        arrange(p_value) %>%
        mutate(p_value = cell_spec(round(p_value,4),"html",color = ifelse(p_value > 0.05,"black","lightgreen"))) %>%
        column_to_rownames("names") %>%
        t %>%
        kable(format = "html", escape = F) %>%
        kable_styling(full_width = F)


data %>%
        dplyr::select(cutc,meat_cat, smoker, fruits,rice,salad,vegetables,full_grain,eggs) %>%
        melt(id="cutc") %>%
         ggplot(aes(x=as.factor(value),y=cutc,fill=as.factor(value)),data=.) +
        geom_boxplot(binaxis = "y", stackdir = "center") +
        scale_y_log10() +
        facet_grid(~variable,scales="free") 

```

## Univariate associations with cnta
```{r echo=FALSE, message=FALSE, warning=FALSE}
data %>%
        dplyr::select(meat_cat, smoker, fruits,rice,salad,vegetables,full_grain,eggs) %>%
        map(~lm(log(data$cnta) ~ .x,data=data)) %>%
        map(summary) %>%
        map(coef) %>%
        map_dbl(function(x) x[[2,4]]) %>%
        tidy %>%
        rename(p_value = x) %>%
        arrange(p_value) %>%
        mutate(p_value = cell_spec(round(p_value,4),"html",color = ifelse(p_value > 0.05,"black","lightgreen"))) %>%
        column_to_rownames("names") %>%
        t %>%
        kable(format = "html", escape = F) %>%
        kable_styling(full_width = F)

data %>%
        dplyr::select(cnta,meat_cat, smoker, fruits,rice,salad,vegetables,full_grain,eggs) %>%
        melt(id="cnta") %>%
        ggplot(aes(x=as.factor(value),y=cnta,fill=as.factor(value)),data=.) +
        geom_boxplot(binaxis = "y", stackdir = "center") +
        scale_y_log10() +
        facet_grid(~variable,scales="free")

```


## Univariate associations with grdh
```{r echo=FALSE, message=FALSE, warning=FALSE}
data %>%
        dplyr::select(meat_cat, smoker, fruits,rice,salad,vegetables,full_grain,eggs) %>%
        map(~lm(log(data$grdh) ~ .x,data=data)) %>%
        map(summary) %>%
        map(coef) %>%
        map_dbl(function(x) x[[2,4]]) %>%
        tidy %>%
        rename(p_value = x) %>%
        arrange(p_value) %>%
        mutate(p_value = cell_spec(round(p_value,4),"html",color = ifelse(p_value > 0.05,"black","lightgreen"))) %>%
        column_to_rownames("names") %>%
        t %>%
        kable(format = "html", escape = F) %>%
        kable_styling(full_width = F)

data %>%
        dplyr::select(grdh,meat_cat, smoker, fruits,rice,salad,vegetables,full_grain,eggs) %>%
        melt(id="grdh") %>%
        ggplot(aes(x=as.factor(value),y=grdh,fill=as.factor(value)),data=.) +
        geom_boxplot(binaxis = "y", stackdir = "center") +
        scale_y_log10() +
        facet_grid(~variable,scales="free")

```


## Univariate associations with principal component 2 - all microbiota
```{r echo=FALSE, message=FALSE, warning=FALSE}
data %>%
        dplyr::select(meat_cat, smoker, fruits,rice,salad,vegetables,full_grain,eggs) %>%
        map(~lm(PC2 ~ .x,data=data)) %>%
        map(summary) %>%
        map(coef) %>%
        map_dbl(function(x) x[[2,4]]) %>%
        tidy %>%
        rename(p_value = x) %>%
        arrange(p_value) %>%
        mutate(p_value = cell_spec(round(p_value,4),"html",color = ifelse(p_value > 0.05,"black","lightgreen"))) %>%
        column_to_rownames("names") %>%
        t %>%
        kable(format = "html", escape = F) %>%
        kable_styling(full_width = F)

data %>%
        dplyr::select(PC2,meat_cat, smoker, fruits,rice,salad,vegetables,full_grain,eggs) %>%
        melt(id="PC2") %>%
        ggplot(aes(x=as.factor(value),y=PC2,fill=as.factor(value)),data=.) +
        geom_boxplot(binaxis = "y", stackdir = "center") +
        scale_y_log10() +
        facet_grid(~variable,scales="free")
```


```{r}
library(table1)

table1(~ factor(meat_cat) + factor(smoker) + fruits + rice + salad + vegetables + full_grain + eggs | as.factor(gender),data=data)

```

##
```{r}
data %>%
  ggplot(aes(x=as.factor(smoker),y=age,fill=as.factor(smoker)),data=.) +
  geom_boxplot()

fit <- lm(log(cutc)~age+smoker,data=data)
summary(fit)
```


```{r}
fit1 <- lm(log(cutc)~fruits+vegetables+gender+age,data=data)
anova(fit1)

tab_model(fit1)

plot(data$fruits,data$age)

data <- data %>% mutate(gesund = (fruits+vegetables)/2)

fit1 <- lm(log(cutc)~gesund+meat_cat+gender,data=data)
anova(fit1)


tab_model(fit1)


```










